### 2020-06-26 latex/vnhcs.tex
Finished a base draft of simply actuated mech systems

Most mechanical systems are not solely a collection of point masses which can 
freely move about three-dimensional space; the masses are usually restricted
physically by some equations of constraint (EOC). 
Rather than representing the system by one vector in \(\R^{3N}\) and \(k\) EOC, 
it is often feasible to represent the position of the system using \(n\)
independent scalar-valued variables \(q_1,\ldots,q_n\) called 
\textit{generalized coordinates}, where \(n = 3N - k\) is the number of
\textit{degrees of freedom} (DOF) of the system.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian
(\ref{eqn:lagrangian-general})
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^m\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times m}\)
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times m}\) given by
\[
    B(q) =\begin{bmatrix}
        - & B_1(q) & - \\
          & \vdots & \\
        - & B_n(q) & -
    \end{bmatrix}
\]
is the \textit{input matrix} for the system. If \(m < n\), we say the system is
\textit{underactuated} with degree of underactuation \(n - m\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations
(\ref{eqn:el-eqns-general}) reduce to (\ref{eqn:el-eqns}),
\begin{equation}\label{eqn:el-eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{[D]_{i,j}}{q_k}  +
     \pdiff{[D]_{i,k}}{q_j} -
     \pdiff{[D]_{k,j}}{q_i}\right)\dot{q}_k
\]

% ---------- Hamiltonian Mechanics ---------- %
\subsection{Hamiltonian Mechanics}
% TODO: Describe how to get the Hamiltonian from the lagrangian, and show that
% you can write the hamiltonian directly in certain conditions of the system
Hamiltonian mechanics provides an equivalent representation of the EOM
by converting the \(n\) second-order ODEs generated by Lagrangian mechanics into
\(2n\) first-order ODEs.

To do this, we first define the \textit{conjugate of momentum to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) = \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we will write \(p = (p_1, \ldots, p_n)\) and say that \(p\) is
the \textit{conjugate of momenta to \(q\)} Note that each \(p_i\) is a linear
function of \(\dot{q}\), so one can typically solve for \(\dot{q}(q,p)\) by
inverting the expressions from (\ref{eqn:p-i}). The combined vector \((q,p)\) is
called the \textit{phase} of the system.

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \textbf{SOURCE}:
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \pdiff{\mathcal{H}}{p} \\
        \dot{p} = -\pdiff{\mathcal{H}}{q} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times m}\) is the input matrix and 
\(\tau \in \R^m\) is the vector of generalized input forces.

If the Lagrangian is given as in (\ref{eqn:lagrangian}), the conjugate of
momenta is computed explicitly by
\[
    p = D(q)\dot{q}
\]
which gives the Hamiltonian (\ref{eqn:hamiltonian})
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2} p\tpose D^{-1}(q) p + P(q)
\end{equation}
and the EOM from (\ref{eqn:hamiltonian-eom-general}) reduce to 
(\ref{eqn:hamiltonian-eom}).
\begin{equation}\label{eqn:hamiltonian-eom}
    \begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p_i} = -\frac{1}{2} p\tpose \pdiff{D^{1}}{q_i}(q) p -
        \pdiff{P}{q_i}(q) + B_i(q) \tau \\
    \end{cases}
\end{equation}

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a mechanical system with phase coordinates \((q,p)\) and Hamiltonian
(\ref{eqn:hamiltonian}), it is not in general obvious how the input forces
\(\tau\) are affecting the conjugate of momenta since they are scaled by
\(B(q)\), which may be quite complicated. Even if \(B(q)\) is constant, multiple
forces may affect each \(p_i\). This section will demonstrate a useful change of
coordinates which results in a simpler input matrix.

Throughout the rest of this thesis, we will make the following assumptions.
\begin{assm}\label{assm:B-const}
    The input matrix \(B : \mathcal{Q} \rightarrow \R^{n\times k}\) is constant
    and full rank.
\end{assm}
\begin{assm}\label{assm:B-orthogonal}
    The input matrix is left-semi-orthogonal, whereby \(B\tpose B = \Id{k}\).
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a right-semi-orthogonal matrix 
    \(B^\perp \in \R^{(n-k)\times n}\) for \(B\)
    which is a left-annihilator for \(B\). That is,
    \(B^\perp(B^\perp)\tpose = \Id{(n-k)}\) and
    \(B^\perp B = \Zmat{(n-k) \times k}\) 
\end{assm}

Observe that Assumption \ref{assm:B-orthogonal} can be taken without loss of
generality (WLOG). If \(B\) is a constant full-rank matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{n \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 & 0 \\
        0 & \sigma_2 & \cdots & 0 & 0 \\
        \vdots & & \ddots &  & \vdots \\
        0 & 0 & \cdots & 0 & \sigma_k \\
          & - & \Zmat{(n-k)\times k} & -  &\\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\).
\textbf{SOURCE}
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau}\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).

If \(k = (n-1)\) then if there is a left annihilator \(A\) for
\(B\), then \(B^\perp = A/\norm{A}\) satisfies Assumption
\ref{assm:B-perp}.

To perform our coordinate transformation, we define a transformation matrix
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since both \(B^\perp\) and \(B\tpose\) are right-semi-orthogonal, 
\(\mathbf{B}\) is orthogonal. That is,
\(\mathbf{B}\mathbf{B}\tpose = \Id{n}\) which means
\(\mathbf{B}^{-1} = \mathbf{B}\tpose\).

\begin{defn}
    We call \(q_u = B^\perp q\) the \textbf{unactuated coordinates} and 
    \(q_a = B\tpose\) the \textbf{actuated coordinates}, which are obtained
    through the transformation
    \[
        \begin{bmatrix} q_u \\ q_a \end{bmatrix} = \mathbf{B}q
    \]
\end{defn}

>>>>

Most mechanical systems are not solely a collection of point masses which can 
freely move about three-dimensional space; the masses are usually restricted
physically by some equations of constraint (EOC). 
Rather than representing the system by one vector in \(\R^{3N}\) and \(k\) EOC, 
it is often feasible to represent the position of the system using \(n\)
independent scalar-valued variables \(q_1,\ldots,q_n\) called 
\textit{generalized coordinates}, where \(n = 3N - k\) is the number of
\textit{degrees of freedom} (DOF) of the system \cite{greenwood_dynamics}.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}:
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^k\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times k}\)
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = [B_i(q)]
\]
is called the \textit{input matrix} for the system. 
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \(n - k\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations reduce to (\ref{eqn:el-eqns}),
\begin{equation}\label{eqn:el-eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{D_{i,j}}{q_k}  +
     \pdiff{D_{i,k}}{q_j} -
     \pdiff{D_{k,j}}{q_i}\right)\dot{q}_k
\]

% ---------- Hamiltonian Mechanics ---------- %
\subsection{Hamiltonian Mechanics}
Hamiltonian mechanics provides an equivalent representation of the EOM
by converting the \(n\) second-order ODEs generated by Lagrangian mechanics into
\(2n\) first-order ODEs.

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) = \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we will write \(p = (p_1, \ldots, p_n)\) and say that \(p\) is
the \textit{conjugate of momenta to \(q\)} Note that each \(p_i\) is a linear
function of \(\dot{q}\), so one can typically solve for \(\dot{q}(q,p)\) by
inverting the expressions from (\ref{eqn:p-i}). The combined vector \((q,p)\) is
called the \textit{phase} of the system.

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \cite{landau_mechanics}:
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations called \textit{Hamilton's equations}
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times k}\) is the input matrix and 
\(\tau \in \R^k\) is the vector of generalized input forces. A pair of
coordinates \((q,p)\) which satisfy Hamilton's equations are said to be 
\textit{canonical coordinates for} \(\mathcal{H}\). a change of coordinates
\((\hat{q}(q,p),\hat{p}(q,p))\) is said to be a \textit{canonical
transformation} if the EOM \((\dot{\hat{q}},\dot{\hat{p}})\) satisfy Hamilton's
equations for some other Hamiltonian \(\hat{H}\).

If the Kinetic energy is quadratic as in (\ref{eqn:lagrangian}), the conjugate of
momenta to \(q\) can be computed explicitly by
\[
    p = D(q)\dot{q}
\]
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p} = -\frac{1}{2} \left[p\tpose \pdiff{D^{-1}}{q_i}(q) p\right] 
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given an arbitrary Hamiltonian mechanical system (\ref{eqn:hamiltonian}) in
canonical coordinates \((q,p)\), it is not obvious how the input forces \(\tau\)
are affecting the conjugate of momenta \(p_i\). This is because \(\tau\) is
transformed by \(B(q)\), which may be quite complicated.

For this reason, we will define a new class of Hamiltonian systems with simple
input matrices, so that the effect of the input forces on the conjugate of
momenta is obvious. This class of systems will be the backbone for the rest of
the theory developed in this thesis.

\begin{defn}
    A Hamiltonian mechanical system (\ref{eqn:hamiltonian}) is said to be 
    \textit{simply actuated} if the input matrix \(B(q) \in \R^{n \times k}\)
    is always of the form
    \[
        B(q) = \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
\end{defn}

Under the following mild assumptions on the input matrix, we will show that
there exists a canonical transformation under which the system is simply
actuated.

\begin{assm}\label{assm:B-const}
    The input matrix \(B : \mathcal{Q} \rightarrow \R^{n\times k}\) is constant
    and full rank.
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a right-semi-orthogonal matrix 
    \(B^\perp \in \R^{(n-k)\times n}\)
    which is a left-annihilator for \(B\). That is,
    \(B^\perp(B^\perp)\tpose = \Id{(n-k)}\) (right semi-orthogonality) and
    \(B^\perp B = \Zmat{(n-k) \times k}\) (left-annihilation).
\end{assm}

Note that if \(k = (n-1)\), the existence of a left annihilator \(A^0\) for
\(B\) implies \(B^\perp := A^0/\norm{A^0}\) satisfies Assumption
\ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    WLOG, the input matrix \(B\) is left semi-orthogonal. 
    That is, \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant full-rank matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{n \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 & 0 \\
        0 & \sigma_2 & \cdots & 0 & 0 \\
        \vdots & & \ddots &  & \vdots \\
        0 & 0 & \cdots & 0 & \sigma_k \\
          & - & \Zmat{(n-k)\times k} & -  &\\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\).
\textbf{SOURCE}
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau}\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since \(B^\perp\) is a left annihilator or \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal (i.e. it is full-rank and has inverse equal to its transpose):
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
    \Rightarrow
    \mathbf{B}^{-1} = \mathbf{B}\tpose
\]

\begin{defn}
    Define the coordinate transformation 
    \((q_u,q_a) = \mathbf{B}q\), where \(q_u = B^\perp q\)
    and \(q_a = B\tpose q\).
    We call \(q_u\) the \textit{unactuated coordinates} and \(q_a\) the
    \textit{actuated coordinates} of the system. 
    Their corresponding conjugate of momenta are 
    \((p_u,p_a) = \mathbf{B}p\).
\end{defn}

In these new coordinates, the generalized input forces directly affect the
actuated conjugate of momenta. That is, \(\tau_i\) only affects \(p_{a,i}\), 
and all \(p_u\) are unaffected by the input forces. We prove this fact through
the following theorem.

\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the change of coordinates 
    \((Q = (q_u, q_a), P = (p_u,p_a))\) 
    is a canonical transformation of the
    original Hamiltonian system (\ref{eqn:hamiltonian}) into
    the simply actuated Hamiltonian system (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{\mathcal{H}}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2}
                \left[P\tpose \pdiff{\Minv}{Q_i}(Q) P \right] 
                - \nabla_Q V(Q) + 
                \begin{bmatrix}
                    \Zmat{(n-k)\times k} \\
                    \Id{k} \\
                \end{bmatrix} \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    For any constant matrix \(A\), the transformation
    \((Q = Aq, P = Ap)\) satisfies
    \(\pdiff{Q_i}{p_s} = \pdiff{P_i}{q_s} = 0\) for all 
    \(i,s \in \{1,\ldots,n\}\). Hence, the Poisson brackets between the
    new coordinates are zero:
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{s = 1}^n \pdiff{Q_i}{p_s}\pdiff{Q_j}{q_s} - 
        \pdiff{Q_i}{q_s}\pdiff{Q_j}{p_s} = 0 \\
        [P_i,P_j] &:= \sum\limits_{s=1}^n \pdiff{P_i}{p_s}\pdiff{P_j}{q_s} -
        \pdiff{P_i}{q_s}\pdiff{P_j}{p_s} = 0
    \end{align*}
    Since the matrix \(A = \mathbf{B}\) is invertible and orthogonal,
    \(A_i (A\tpose_j)\tpose = A_i (A^{-1}_j)\tpose = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between the new momenta and
    configurations are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{s=1}^n\pdiff{P_i}{p_s}\pdiff{Q_j}{q_s}
        - \pdiff{P_i}{q_s}\pdiff{Q_j}{p_s} \\
                  &= \sum\limits{s=1}^n A_{i,s}A_{j,s} - 0 \\
                  &= \sum\limits{s=1}^n A_{i,s}A\tpose_{s,j} \\
                  &= A_i (A\tpose_j)\tpose \\
                  &= \delta_{i,j}
    \end{align*}

    By (45.10) in \cite{landau_mechanics}, the coordinate change
    \((Q = \mathbf{B}q, P = \mathbf{B}p)\) is a canonical transformation with
    new Hamiltonian 
    \(\hat{H}(Q,P) = \mathcal{H}(\mathbf{B}\tpose Q,\mathbf{B}\tpose P)\).

    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \hat{B} = \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the new Hamiltonian system is simply actuated, as desired.
\end{proof}
